---
title: "PracticalMachineLearning"
author: "Evgieni Gordienko"
date: "December 21, 2014"
output: html_document
---

Practical Machine Learning - Prediction Assignment Writeup
==========================================================

Background and Task

Using devices such as Jawbone Up, Nike FuelBand, and Fitbit it is now possible to collect a l
arge amount of data about personal activity relatively inexpensively. These type of devices are
part of the quantified self movement – a group of enthusiasts who take measurements about 
themselves regularly to improve their health, to find patterns in their behavior, or because they
are tech geeks. One thing that people regularly do is quantify how much of a particular activity
they do, but they rarely quantify how well they do it. In this project, your goal will be to use 
data from accelerometers on the belt, forearm, arm, and dumbell of 6 participants. They were
asked to perform barbell lifts correctly and incorrectly in 5 different ways. More information
is available from the website here: http://groupware.les.inf.puc-rio.br/har (see the section on
the Weight Lifting Exercise Dataset). 

Data sources (copied to local RStudio project folder):

The training data for this project are available here: 
https://d396qusza40orc.cloudfront.net/predmachlearn/pml-training.csv

The test data are available here: 
https://d396qusza40orc.cloudfront.net/predmachlearn/pml-testing.csv

Since training data contains some garbage in form #DIV/0! we have to substitute it with NA.

```{r}
training_data_source <- read.csv("pml-training.csv", na.strings=c("#DIV/0!") )
testing_data_source <- read.csv("pml-testing.csv", na.strings=c("#DIV/0!") )
```

Loading base libraries:

```{r}
library(Hmisc); library(caret);
```

Data preparation and partitioning:

```{r}
set.seed(12345)
inTrain <- createDataPartition(y=training_data_source$classe, p=0.9, list=FALSE)
training <- training_data_source[inTrain,]
testing <- training_data_source[-inTrain,]
```

Cleaning the data - remove near zero covariates:

```{r}
nearZero <- nearZeroVar(training, saveMetrics=TRUE)
```

To find out the list of column names for near zero covariates we use: 

```{r}
nearZero 
```

To speed Up calculations by parallel execution we use:
```{r}
library(doMC)
registerDoMC(cores = 6) 
```
(my box has 8 cores detected)

After filtering near zero covariates out we can train on remaining columns.
We use Random Forest model due to its good ability to estimate the importance of each 
predictor variable in modeling the response variable.
```{r}
library(randomForest)

modFit <- randomForest(classe ~ user_name + pitch_arm + yaw_arm + roll_arm + roll_belt 
+ pitch_belt + yaw_belt + gyros_belt_x + gyros_belt_y + gyros_belt_z + accel_belt_x
+ accel_belt_y + accel_belt_z + magnet_belt_x + magnet_belt_y + magnet_belt_z 
+ gyros_arm_x + gyros_arm_y + gyros_arm_z + accel_arm_x + accel_arm_y + accel_arm_z 
+ magnet_arm_x + magnet_arm_y + magnet_arm_z + roll_dumbbell + pitch_dumbbell
+ yaw_dumbbell, importance = TRUE,data=training, keep.forest=TRUE)

modFit

```

Now verify predictions on testing partition:

```{r}
predictTst<-predict(modFit, testing); confusionMatrix(predictTst, testing$classe);
```

Check accuracy:
```{r}
accuracy <- postResample(predictTst, testing$classe)
accuracy
```

Now Check importance of covariates:

```{r}
varImpPlot(modFit, type=2)
```

roll_belt and yaw_belt are the leaders.

Cross validation and error estimates

```{r}
library(ggplot2)
pred <- predict(modFit, testing); testing$predRight <- pred==testing$classe;
qplot(classe, data=testing, main="Predictions") +  facet_grid(predRight ~ .)
```

Verified that no overfitting.


Validate model on supplied separate testing_data_source:

```{r}
predict_validation<- predict(modFit, testing_data_source)
predict_validation
```

 
FINAL:
put the data into a format ready for the Submission phase of the assignment. 
The prediction should have work as scored 20/20. Use supp;ied by COURSERA function:

```{r}
pml_write_files = function(x){
   n = length(x)
   for(i in 1:n){
     filename = paste0("./submit/problem_id_",i,".txt")
     write.table(x[i],file=filename,quote=FALSE,row.names=FALSE,col.names=FALSE)
   }
 }

```

Run function and produce output

```{r}
pml_write_files(predict_validation)
```

Confirmed after upload of answers in submit folder to Coursera that it is 20/20 scored.
